Viscous Erosion in Porous Media

### 2018-01-31 docs/methods/erosion_methods.tex
worked on intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process that is responsible for a variety
of geophysical and other processes.  For example, wind erosion causes
sand dunes to be  longer on the windward side~\cite{han1969}, and
erosion is responsible for branching structures of streams and
rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is also important in
the biophysical community. Plaque erosion in arteries is key to
understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}.  Another applications is
erosion caused by groundwater flow, where a slow moving fluid erodes
channels, or conduits, in the ground.  This results in a heterogeneous
media that has a large effect on the resulting flow.  In all these
examples, there is a feedback process between the flow and the erodible
bodies---the shape of the eroding body affects the flow, while the flow
changes the shape by fluidizing the eroding body.  This fluid-mechanical
feedback system has been simulated in idealized cases such as a single
eroding two-dimensional body in a uniform
flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. To further develop insight
into the effect of two-dimensional erosion, we consider erosion of a
multiply-connected domain (Figure~\ref{fig:50bodies}), such as the
porous geometry of many groundwater flows.

Because of the scales involved with groundwater flow, we will restrict
our attention to Stokesian dynamics.  For groundwater flow, Darcy's law
is often used, but this assumes a homogeneous geometry which is often
not valid, and the method does not resolve the pore structure.  Instead, we
will use the incompressible Stokes equations to represent the flow, and
the individual pores will be be resolved, resulting in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited as it can resolve complex geometries and the non-negligible
interactions between bodies are resolved. Moreover, high-order
accuracy and computational efficiency are obtained. To link the body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is easily distorted and tangled by the dynamics
and regions of high curvature develop.  We adopt a \thL~framework that
keeps the mesh equidistributed in arclength, and a curvature
penalization term can be added without introducing any non-linear
stiffness.
\todo[inline]{rework this}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study will develop numerical
methods to simulate viscous erosion of multiply-connected
two-dimensional geometries.  By assuming the flow is viscous, we are
assuming the flow is very slow and in small length scales, as is the
case in groundwater flow.  Moreover, by allowing for complex
multiply-connected domains, the effects due to interacting bodies on the
erosion are captured.  We start by authenticating our method on a single
circular body placed symmetrically in a pipe flow.  By artificially
fixing the area of the body as a function of time, we are able to
characterize the long-time behavior of the shape and the resulting flow
and compare these results with analytical results.  Then, we compute the
vorticity of erosion in multiply-connected geometries to better
characterize erosion in porous media.

Because of the low Reynolds number assumption, the flow will be assumed
to be Stokesian.  Then, to create high-fidelity and computationally
efficient numerical methods, two fast and accurate methods will be
employed:
\begin{enumerate}
  \item The fluid equations will be solved with spectral accuracy by
  using an efficient boundary integral equation method. Moreover, the
  rate of erosion will be computed and the rate of erosion will be
  computed with spectral accuracy.
  \item Given a rate of erosion, the shape will be integrated in time
  with second-order accuracy.  All spatial derivatives will be computed
  in Fourier space, resulting in spectral accuracy in space.
\end{enumerate}
We use a quasi-steady assumption so that these two steps can be iterated
to simulate the dynamics.

@@

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  While recent work does allow for
integral equation methods to resolve a corner, we employ a different
method.  First, we add a regularization term to the shear stress to
penalize regions of high curvature.  Second, we apply a Gaussian filter
to smooth the absolute value of the shear stress before the body shape
is updated.

To further visualize the effect of erosion, we compute the vorticity of
the flow (Figure~\ref{fig:50bodies}) which is equal to the shear stress
in the Stokesian limit.  Finally, to evolve the geometry, a second-order
time stepping method that guarantees that the mesh does not become
distorted will be employed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, Several flows can be described
to good accuracy under a two-dimensional approximation, certain
phenomena occur only in three dimensions.  The boundary integral
equation formulation for the fluid solver exists, and the main
challenges of a BIE solver have been addressed by other groups including
surface representation~\cite{yin-bir-zor2006, vee-rah-bir-zor2011},
near-singular integration~\cite{yin-bir-zor2006, kli-tor2016b}, and fast
summation methods~\cite{yin-bir-zor2004, ros-ols2015}.  The more
significant challenge of extending our work to three dimensions is
guaranteeing that the mesh does not become distorted, which can be done,
for example, by using an extension of the two-dimensional {\thL}
formulation~\cite{amb-sie-tlu2013}.

The second limitation is that the eroding bodies are fixed.  Mitchell
and Spagnolie~\cite{mit-spa2016} simulate a single mobile eroding body
in three dimensions have been considered~\cite{mit-spa2016}, but
additional challenges are caused when there are multiple bodies. In
particular, especially in dense suspensions, bodies can become extremely
close to one another, and the interaction between the bodies is
difficult to resolve unless the underlying mesh is refined to levels
that are intractable. In future work we plan to consider mobile eroding
bodies, but include a small force and torque so that unphysical contact
is avoided~\cite{lu-rah-zor2017}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction\label{s:intro}}
Erosion is a fluid-mechanical process that is responsible for a variety
of phenomena in geophysical fluids and other processes.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, and erosion is responsible for branching structures
of streams and rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in the biophysical community. Plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}.  Another application,
which motivates the current work, is erosion caused by groundwater flow.
Here, a slow moving fluid erodes channels in the ground, and this
affects the fluid transport.  In all these examples, there is a feedback
process between the flow and the eroding bodies---the shape of the
bodies affect the flow, while the flow changes the shape by fluidizing
each body.  This fluid-mechanical feedback system has been simulated in
idealized cases such as a single eroding two-dimensional body in a
uniform flow~\cite{moo-ris-chi-zha-she2013} or a single mobile eroding
three-dimensional body~\cite{mit-spa2016}. However, to better understand
the effect of erosion on groundwater flow, multiply-connected domains
must be considered.  Here we simulate two-dimensional erosion, in
multiply-connected domain such as the porous geometry of many
groundwater flows (Figure~\ref{fig:50bodies}).

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater flow,
Darcy's law is often used, but this does not resolve the pore structure
and assumes a homogeneous geometry which is often not valid.  Instead,
we use the incompressible Stokes equations to represent the flow, and
the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a linear diffusive term that can easily be treated implicitly
to avoid stiffness.

In addition to simulating the eroding grains, the flow is further
characterized by computing the vorticity in the flow.  The vorticity is
particularly insightful since it is identical to the shear stress on
no-slip boundaries in the Stokesian limit.  Therefore, by computing the
vorticity, we are able to locate regions where the erosion rates will be
fastest or slowest.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study develops numerical methods
to simulate erosion of multiply-connected two-dimensional geometries.
Since groundwater flow occurs at small spatial and velocity scales, we
assume the flow is viscous (ie.~zero Reynolds number).  By allowing for
complex multiply-connected domains, we capture the effect that
interacting bodies have on erosion, and the grains in the flow are
individually resolved rather than homogenized.  We start by validating
our method on a single circular body placed symmetrically in a pipe
flow.  By artificially fixing the area of the body as a function of
time, we are able to characterize the long-time behavior of the shape
and the resulting flow and compare these results with analytical
results.  Then, we compute the vorticity of erosion in
multiply-connected geometries to better characterize erosion in porous
media.

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid equations
and the erosion rate (shear stress) are efficiently solved with spectral
accuracy by using a fast multipole accelerated boundary integral
equation method.  Given the rate of erosion, the shape is integrated in
time with second-order accuracy.  All spatial derivatives are be
computed in Fourier space, resulting in spectral accuracy in space.
Then, to further visualize the effect of erosion, we compute the
vorticity of the flow (Figure~\ref{fig:50bodies}) to visualize regions
of high and low erosion.

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  While recent work does allow for
integral equation methods to resolve corners, we employ a different
method.  First, we add a regularization term to the shear stress to
penalize regions of high curvature.  Second, we apply a Gaussian filter
to smooth the absolute value of the shear stress before the body shape
is updated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Limitations} There are two main limitations of the current
work.  First, the flow is assumed to be two-dimensional.  However, by
assuming that the flow is two-dimensional, we are able to simulate the
erosion of more bodies and with higher resolution than would be
available in three dimensions.  Moreover, several flows can be described
to good accuracy under a two-dimensional approximation, certain
phenomena occur only in three dimensions.  Extending the current work to
three dimensions is possible with a boundary integral equation
formulation.  In particular, the fluid solvers exist~\cite{mal-bir2015},
and the main challenges of a BIE solver have been addressed by other
groups including surface representation~\cite{yin-bir-zor2006,
vee-rah-bir-zor2011}, near-singular integration~\cite{yin-bir-zor2006,
kli-tor2016b}, and fast summation methods~\cite{yin-bir-zor2004,
ros-ols2015}.  The more significant challenge of extending our work to
three dimensions is guaranteeing that the mesh does not become
distorted, which can be done, for example, by using an extension of the
two-dimensional {\thL} formulation~\cite{amb-sie-tlu2013}.

The second limitation is that the eroding bodies are fixed.  Mitchell
and Spagnolie~\cite{mit-spa2016} simulate a single mobile eroding body
in three dimensions~\cite{mit-spa2016}, but additional challenges are
caused when there are multiple bodies. In particular, especially in
dense suspensions, bodies can become extremely close to one another, and
the interaction between the bodies is difficult to resolve unless the
underlying mesh is refined to levels that are intractable. In future
work we plan to consider mobile eroding bodies, but include a small
force and torque so that unphysical contact is
avoided~\cite{lu-rah-zor2017}.

Mitchell and Spagnolie~\cite{mit-spa2016} formulate the traction jump as
a DLP using Lorentz Reciprocal Theorem so that additional layer
potentials have to be evaluated.  That is, they never find the density
function.

@@
